$K$-corrections and filter transformations \\ from the restframe ultraviolet to the near infrared

### 2006-05-16 docs/paper/kcorrect.tex
adjust
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@552 75dff457-8751-0410-ab1a-cd701654ce6b

\newcommand{\XX}{X}
\newcommand{\XXt}{\tilde{X}}
\newcommand{\XXh}{\hat{X}}
\newcommand{\Xij}{X_{ij}}
\newcommand{\Xijh}{\hat{X}_{ij}}
\renewcommand{\AA}{A}
\newcommand{\AAt}{\tilde{A}}
\newcommand{\WW}{W}
\newcommand{\WWt}{\tilde{W}}
\newcommand{\Wik}{W_{ik}}
\newcommand{\HH}{H}
\newcommand{\HHt}{\tilde{H}}
\newcommand{\Hkj}{H_{kj}}
\newcommand{\BB}{B}
\newcommand{\BBt}{\tilde{B}}
\newcommand{\MM}{M}
\newcommand{\MMt}{\tilde{M}}
\newcommand{\Aik}{A_{ik}}
\newcommand{\Bkr}{B_{kr}}
\newcommand{\Mrj}{M_{rj}}
\newcommand{\sij}{\sigma_{ij}}

@@

\author{
Michael R. Blanton\altaffilmark{\ref{NYU}} and 
Sam Roweis
%Tamas Budavari\altaffilmark{\ref{JHU}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%J.~Brinkmann\altaffilmark{\ref{APO}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Mamoru Doi\altaffilmark{\ref{Tokyo}},
%Daniel Eisenstein\altaffilmark{\ref{Arizona}},
%Masataka Fukugita\altaffilmark{\ref{CosmicRay},\ref{IAS}},
%James E. Gunn\altaffilmark{\ref{Princeton}},
%David W. Hogg\altaffilmark{\ref{NYU}}, and
%David J. Schlegel\altaffilmark{\ref{Princeton}}
%Julianne Dalcanton\altaffilmark{\ref{UW}},
%Jon Loveday\altaffilmark{\ref{Sussex}},
%Michael A. Strauss\altaffilmark{\ref{Princeton}},
%Mark SubbaRao\altaffilmark{\ref{Chicago}},
%David H. Weinberg\altaffilmark{\ref{Ohio}},
%John E. Anderson, Jr.\altaffilmark{\ref{Fermilab}},
%James Annis\altaffilmark{\ref{Fermilab}},
%Neta A. Bahcall\altaffilmark{\ref{Princeton}},
%Mariangela Bernardi\altaffilmark{\ref{Chicago}},
%Robert J. Brunner\altaffilmark{\ref{Caltech}},
%Scott Burles\altaffilmark{\ref{Fermilab}},
%Larry Carey\altaffilmark{\ref{UW}},
%Francisco J. Castander\altaffilmark{\ref{Chicago}, \ref{Pyrenees}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Douglas Finkbeiner\altaffilmark{\ref{Berkeley}},
%Scott Friedman\altaffilmark{\ref{JHU}},
%Joshua A. Frieman\altaffilmark{\ref{Fermilab}},
%G. S. Hennessy\altaffilmark{\ref{USNO}},
%Robert B. Hindsley\altaffilmark{\ref{USNO}},
%Takashi Ichikawa\altaffilmark{\ref{Tokyo}},
%\v{Z}eljko Ivezi\'{c}\altaffilmark{\ref{Princeton}},
%Stephen Kent\altaffilmark{\ref{Fermilab}},
%G. R.~Knapp\altaffilmark{\ref{Princeton}},
%D. Q.~Lamb\altaffilmark{\ref{Chicago}},
%R. French Leger\altaffilmark{\ref{UW}},
%Daniel C. Long\altaffilmark{\ref{APO}},
%Robert H. Lupton\altaffilmark{\ref{Princeton}},
%Timothy A.~McKay\altaffilmark{\ref{Michigan}},
%Avery Meiksin\altaffilmark{\ref{Edinburgh}},
%Aronne Merelli\altaffilmark{\ref{Caltech}},
%Jeffrey A. Munn\altaffilmark{\ref{USNO}},
%Vijay Narayanan\altaffilmark{\ref{Princeton}},
%Matt Newcomb\altaffilmark{\ref{CarnegieMellon}},
%R. C. Nichol\altaffilmark{\ref{CarnegieMellon}},
%Sadanori Okamura\altaffilmark{\ref{Tokyo}},
%Russell Owen\altaffilmark{\ref{UW}},
%Jeffrey R.~Pier\altaffilmark{\ref{USNO}},
%Adrian Pope\altaffilmark{\ref{JHU}},
%Marc Postman\altaffilmark{\ref{STScI}},
%Thomas Quinn\altaffilmark{\ref{UW}},
%Constance M. Rockosi\altaffilmark{\ref{Chicago}},
%Donald P. Schneider\altaffilmark{\ref{PennState}}, 
%Kazuhiro Shimasaku\altaffilmark{\ref{Tokyo}},
%Walter A. Siegmund\altaffilmark{\ref{UW}},
%Stephen Smee\altaffilmark{\ref{Maryland}},
%Yehuda Snir\altaffilmark{\ref{CarnegieMellon}},
%Chris Stoughton\altaffilmark{\ref{Fermilab}},
%Christopher Stubbs\altaffilmark{\ref{UW}},
%Alexander S.~Szalay\altaffilmark{\ref{JHU}},
%Gyula P.~Szokoly\altaffilmark{\ref{Potsdam}},
%Aniruddha R.~Thakar\altaffilmark{\ref{JHU}},
%Christy Tremonti\altaffilmark{\ref{JHU}},
%Douglas L. Tucker\altaffilmark{\ref{Fermilab}},
%Alan Uomoto\altaffilmark{\ref{JHU}},
%Dan vanden Berk\altaffilmark{\ref{Fermilab}},
%Michael S. Vogeley\altaffilmark{\ref{Drexel}},
%Patrick Waddell\altaffilmark{\ref{UW}},
%Brian Yanny\altaffilmark{\ref{Fermilab}},
%Naoki Yasuda\altaffilmark{\ref{NAOJ}},
%and Donald G.~York\altaffilmark{\ref{Chicago}}
}

\altaffiltext{1}{Based on observations obtained with the
Sloan Digital Sky Survey} 
\setcounter{address}{2}

@@

New York University, Department of Physics, 4 Washington Place, New

@@

@@

The templates are designed to work well for a wide range of data sets.
We constrain them with photometry and spectroscopy of the Sloan
Digital Sky Survey (SDSS; \citealt{york00a}) the Galaxy Evolution
Explorer (GALEX; \citealt{martin05a}) in the ultraviolet, and the
Two-Micron All Sky Survey (2MASS; \citealt{skrutskie97a}) in the near
infrared (NIR). In addition, at higher redshifts, we use constraints
from the Deep Extragalactic Evolutionary Probe 2 (DEEP2;
\citealt{davis03a, faber03a}) and the Great Observatories Origins Deep
Survey (GOODS;
\citealt{giavalisco04a}).  These and other data sets provide a huge set of
information about galaxy colors and spectra which we can use to help
understand their star-formation histories.

We note that the methods used here are generally useful. Using these
methods to develop templates has the advantage that one can easily
search a large space of models in order to find the few templates
necessary in order to explain a given set of data.

We also release the templates in electronic form as well as an
implementation of the methods used to fit the templates to data. This
software {\tt kcorrect v4\_1} is distributed on the World Wide
Web{\footnote {\tt http://cosmo.nyu.edu/blanton/kcorrect/}}. It
consists of a core C library which performs most of the complex and
computationally intensive tasks, plus an IDL library that provides a
high-level interface.  This software is an update of two major earlier
releases ({\tt v1\_16}; \citealt{blanton03b}; {\tt v3\_2}). The
improvement over the latest version is twofold. First, we have
improved the templates such that they successfully fit galaxies in the
restframe UV, as observed by GALEX, DEEP2, and GOODS. Second, because
the templates are now completely model-based, the fits have a physical
interpretation in terms of a star-formation history. The IDL library
also has a number of new useful functions.

In Section \ref{algorithm}, we describe how to find suitable
templates, given a set of models and a set of data. We also describe
the data and models used here. In Section \ref{results}, we show the
results: the best-fit templates and how well they fit the data.

Where necessary, we have assumed cosmological parameters $\Omega_0 =
0.3$, $\Omega_\Lambda = 0.7$, and $H_0 = 100$ $h$ km s$^{-1}$
Mpc$^{-1}$ (with $h=1$). All magnitudes are (unless otherwise noted)
AB-relative.  Except where noted, I will always be referring to the
version of the {\tt kcorrect} software labeled {\tt
v4\_1\_4}.

@@

From the spectroscopic observations we know that galaxy spectra reside
in a low-dimensional locus. Principal Component Analysis (PCA) of
galaxy spectra, introduced by \citet{connolly95b} and applied many
times since then, demonstrates that most of the variance in the
distribution of galaxies in spectral space can be explained using a
few templates. This means that, in the linear space of all possible
spectra, galaxies exist in only a small locus. Therefore, even with
very heterogeneous data, we should be able to determine the properties
of this locus.

Here we present an approach to combining the heterogeneous data
described above in order to determine the properties of the locus of
galaxy spectra. Rather than taking the model-free approach used by
PCA, we here restrict the space of possible spectra to those predicted
from the high resolution stellar population synthesis model of
\citet{bruzual03a} and the nebular emission line models of \citet{kewley01a}.
This approach both constrains the problem appropriately and yields a
natural theoretical interpretation of the results in terms of
star-formation histories.

In a nutshell, our algorithm does the following. Given the
observations (and uncertainties of those observations) available for
each galaxy, it finds the nonnegative linear combination of $N$
template star-formation histories which best predict those
observations in the $\chi^2$ sense. Given the entire set of galaxy
observations available, it also fits for those template star-formation
histories. The technical name for this algorithm is Nonnegative Matrix
Factorization (NMF). Note that this problem is not the same as the
well-known nonnegative least squares problem, which would fit for the
best-fit nonnegative linear combination of templates, but not for the
form of the templates themselves.

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, it has a natural physical
interpretation associated with that spectral subspace, which is the
corresponding subspace of all possible star-formation
histories. Second, it naturally handles data uncertainties and missing
data, which allows it to ignore that variation which is due purely to
statistical errors.  Third, it handles the complications of observing
galaxy spectra photometrically using broad-band filters of galaxies at
varying redshifts, whereas standard PCA requires a constant grid of
observations in rest-frame wavelength space.

In this section, we describe how to take the astrophysical models and
the astronomical observations, and to pose the problem in the form
needed for the purposes of NMF. Section \ref{models} describes the
basis set of models, Section {\bf redo} In Appendix \ref{nmf} we describe
how to solve to NMF problem.

@@

Let us refer to these 485 basis templates as $M_{\lambda,
j}(\lambda)$, expressed in units of ergs s$^{-1}$ \AA$^{-1}$.

We seek to reduce this full basis space to a subspace in which
galaxies actually exist. In particular, we seek five templates
$F_{\lambda, i}(\lambda)$ built from nonnegative combinations of the
original basis set of $N$ templates:
\begin{equation}
F_{\lambda, i}(\lambda) = \sum_j b_{ij} M_{\lambda, j}(\lambda),
\end{equation}
in units of ergs s$^{-1}$ \AA$^{-1}$. In principle, we could seek any
number of templates; from our experiments, we have found that five
turns out to be a number large enough to explain the data we use
here. For each galaxy $k$ we want our model
${\hat{F}_\lambda}(\lambda)$ for their spectrum to be a nonegative sum
of these five templates:
\begin{equation}
{\hat{F}}_{\lambda,_k}(\lambda) = \sum_i a_{ki} F_{\lambda,i}(\lambda).
\end{equation}

\subsection{The data}

@@

\subsection{Comparing data and models}

@@

For the spectra, we take the observed spectrum
$f_{\lambda,k}(\lambda)$ (in ergs cm$^{-2}$ s$^{-1}$ \AA$^{-1}$) of
each galaxy $k$ at redshift $z$ (corresponding to a luminosity
distance $d_L(z)$ in cm; see \citealt{hogg99a}) and calculate the
restframe luminosity per unit wavelength:
\begin{equation}
{{F}}_{\lambda, k}(\lambda) = 
{{f}}_{\lambda, k}[\lambda (1+z)](1+z) (4 \pi d_L^2),
\end{equation}
That is, the spectrum is shifted due to the redshift, while the
integral of the numerator over wavelength (the total luminosity) is
constant with redshift, and the total flux is related to the total
luminosity by the inverse square law.  In addition, we smooth each
spectrum by such an amount that, given its estimated velocity
dispersion, its total velocity dispersion after smoothing is 300 km
s$^{-1}$ (there are a small number of galaxies with larger velocity
dispersions, which we leave unchanged).

@@

If one has a spectrum for a galaxy the
expression for the contribution to $\chi^2$ from each wavelength
$\lambda_l$ in the spectrum is then quite simple:
\begin{eqnarray}
\label{chi2spec}
\chi_{kl}^2 &=& \left[\frac{F_{\lambda,k}(\lambda_l) -
{\hat{F}}_{\lambda,k}(\lambda_l)} 
{\sigma^2_k(\lambda_l)}\right]^2 \cr
&=&
\left[\frac{F_{\lambda,k}(\lambda_l) -
\sum_{ij} a_{ki} b_{ij} M_{jl}
}{\sigma^2_k(\lambda_l)}\right]^2. 
\end{eqnarray}

@@

Here we express the flux for each galaxy $k$ in units of AB maggies
$\mu_p$, which are defined as what one would measure in bandpass $p$
relative to the AB standard source. For example, if we transform our
spectral energy density basis function $M_{\lambda, j}(\lambda)$ to a flux
density $m_{\lambda, j}(\lambda)$, we can calculate the contribution of
that basis function to the predicted maggies as:
\begin{equation}
\label{maggies}
\mu_{jp} = 
\frac{\int_0^{\infty} d\lambda \lambda R_p(\lambda) {m}_{\lambda, j}(\lambda)}
{\int_0^\infty d\lambda \lambda R_p(\lambda) f_{\lambda,
\mathrm{AB}}(\lambda) }
\end{equation}
Here, the response function $R_p(\lambda)$ is proportional to the
contribution to the detector signal of a photon with wavelength
$\lambda$ entering the Earth's atmosphere (or entering the telescope
for a space telescope). The AB standard source is $f_{\lambda,
\mathrm{AB}}(\lambda) d\lambda = f_{\nu,
\mathrm{AB}} (\nu) d\nu$ and $f_{\nu, \mathrm{AB}}=3631$ Jy $= 3.631 \times
10^{-20}$ erg s$^{-1}$ cm$^{-2}$ Hz$^{-1}$. Of course maggies $\mu$
are related to magnitudes $m$ as:
\begin{equation}
m = -2.5 \log_{10} \mu, 
\end{equation}
such that the AB standard source would (if it existed) have $\mu=1$
and $m=0$ for all bandpasses.

In this context it is worth noting that many authors
(e.g. \citealt{bessell90a}) tabulate the contribution to the detector
signal per {\it unit of energy} in photons of wavelength $\lambda$
instead of per {\it photon} with wavelength $\lambda$.  We will refer here
to that quantity as $R_p'(\lambda)$, though in the literature it is
often referred to without a prime (and without any explicit
definition!). Clearly $R_p(\lambda) \propto R_p'(\lambda)/\lambda$,
since the higher the frequency of the photon, at a fixed response per
unit energy there is a higher response per photon. With this
substitution, one can reexpress Equation
\ref{maggies} appropriately in terms of $R_p'(\lambda)$.  
Generally, though not universally, authors tabulate $R_p'(\lambda)$
for bandpasses whose standards were originally calibrated using
energy-counting devices rather than photon-counting devices. However,
from the point of view of the analysis of the observations it is
irrelevant what the devices used for the standards and the
observations are, as long as one calculates $R_p(\lambda)$ and uses
Equation \ref{maggies}.

@@

The matrices $\eta_{jp}$ (related to the broad-band fluxes) and
$M_{jl}$ (related to the spectra) are totally fixed, and we combine
them into a single matrix $M_{jn}$ and the indices $p$ and $l$ into a
single index $n$, to handle both the broad-band fluxes and spectra
simultaneously. We can similarly combine our observations $\mu_{kp}$
and $F_{\lambda, k}(\lambda_l)$ and their uncertainties $\sigma_{kp}$
and $\sigma_{k}(\lambda_l)$ into vectors $x_{kn}$ and $\sigma_{kn}$.
Then we can combine the Equations \ref{chi2spec} and
\ref{chi2photo} into a single equation:
\begin{equation}
\label{chi2}
\chi^2 = \sum_{kn} \left[ 
\frac{x_{kn} - \sum_{ij} a_{ki} b_{ij} M_{jn}}
{\sigma_{kn}} \right]^2
\end{equation}
There is a simple method that we describe in Appendix \ref{nmf} called
nonnegative matrix factorization (NMF) to iterate to the nonnegative
$a_{ki}$ and $b_{ij}$ which minimize Equation \ref{chi2}. The basic
method is implemented in a public piece of code named {\tt
NMF\_SPARSE} in the {\tt idlutils} distribution of IDL
utilities.\footnote{\tt http://skymaps.info}. As the name implies, our
implementation takes advantage of the fact that many of the matrix
operations are on very sparse matrices (for example, for each galaxy
with photometric data there are {\it no} spectroscopic data points and
{\it only} photometric data at a single redshift).

{\bf nonconvexity; not reproducible}

>>>>

\newcommand{\XX}{X}
\newcommand{\XXt}{\tilde{X}}
\newcommand{\XXh}{\hat{X}}
\newcommand{\Xkn}{X_{kn}}
\newcommand{\Xknh}{\hat{X}_{kn}}
\renewcommand{\AA}{A}
\newcommand{\AAt}{\tilde{A}}
\newcommand{\WW}{W}
\newcommand{\WWt}{\tilde{W}}
\newcommand{\Wki}{W_{ki}}
\newcommand{\HH}{H}
\newcommand{\HHt}{\tilde{H}}
\newcommand{\Hin}{H_{in}}
\newcommand{\BB}{B}
\newcommand{\BBt}{\tilde{B}}
\newcommand{\MM}{M}
\newcommand{\MMt}{\tilde{M}}
\newcommand{\Aki}{A_{ki}}
\newcommand{\Bij}{B_{ij}}
\newcommand{\Mjn}{M_{jn}}
\newcommand{\skn}{\sigma_{kn}}

@@

\author{
Michael R. Blanton\altaffilmark{\ref{NYU}} and 
Sam Roweis\altaffilmark{\ref{UToronto}}
%Tamas Budavari\altaffilmark{\ref{JHU}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%J.~Brinkmann\altaffilmark{\ref{APO}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Mamoru Doi\altaffilmark{\ref{Tokyo}},
%Daniel Eisenstein\altaffilmark{\ref{Arizona}},
%Masataka Fukugita\altaffilmark{\ref{CosmicRay},\ref{IAS}},
%James E. Gunn\altaffilmark{\ref{Princeton}},
%David W. Hogg\altaffilmark{\ref{NYU}}, and
%David J. Schlegel\altaffilmark{\ref{Princeton}}
%Julianne Dalcanton\altaffilmark{\ref{UW}},
%Jon Loveday\altaffilmark{\ref{Sussex}},
%Michael A. Strauss\altaffilmark{\ref{Princeton}},
%Mark SubbaRao\altaffilmark{\ref{Chicago}},
%David H. Weinberg\altaffilmark{\ref{Ohio}},
%John E. Anderson, Jr.\altaffilmark{\ref{Fermilab}},
%James Annis\altaffilmark{\ref{Fermilab}},
%Neta A. Bahcall\altaffilmark{\ref{Princeton}},
%Mariangela Bernardi\altaffilmark{\ref{Chicago}},
%Robert J. Brunner\altaffilmark{\ref{Caltech}},
%Scott Burles\altaffilmark{\ref{Fermilab}},
%Larry Carey\altaffilmark{\ref{UW}},
%Francisco J. Castander\altaffilmark{\ref{Chicago}, \ref{Pyrenees}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Douglas Finkbeiner\altaffilmark{\ref{Berkeley}},
%Scott Friedman\altaffilmark{\ref{JHU}},
%Joshua A. Frieman\altaffilmark{\ref{Fermilab}},
%G. S. Hennessy\altaffilmark{\ref{USNO}},
%Robert B. Hindsley\altaffilmark{\ref{USNO}},
%Takashi Ichikawa\altaffilmark{\ref{Tokyo}},
%\v{Z}eljko Ivezi\'{c}\altaffilmark{\ref{Princeton}},
%Stephen Kent\altaffilmark{\ref{Fermilab}},
%G. R.~Knapp\altaffilmark{\ref{Princeton}},
%D. Q.~Lamb\altaffilmark{\ref{Chicago}},
%R. French Leger\altaffilmark{\ref{UW}},
%Daniel C. Long\altaffilmark{\ref{APO}},
%Robert H. Lupton\altaffilmark{\ref{Princeton}},
%Timothy A.~McKay\altaffilmark{\ref{Michigan}},
%Avery Meiksin\altaffilmark{\ref{Edinburgh}},
%Aronne Merelli\altaffilmark{\ref{Caltech}},
%Jeffrey A. Munn\altaffilmark{\ref{USNO}},
%Vijay Narayanan\altaffilmark{\ref{Princeton}},
%Matt Newcomb\altaffilmark{\ref{CarnegieMellon}},
%R. C. Nichol\altaffilmark{\ref{CarnegieMellon}},
%Sadanori Okamura\altaffilmark{\ref{Tokyo}},
%Russell Owen\altaffilmark{\ref{UW}},
%Jeffrey R.~Pier\altaffilmark{\ref{USNO}},
%Adrian Pope\altaffilmark{\ref{JHU}},
%Marc Postman\altaffilmark{\ref{STScI}},
%Thomas Quinn\altaffilmark{\ref{UW}},
%Constance M. Rockosi\altaffilmark{\ref{Chicago}},
%Donald P. Schneider\altaffilmark{\ref{PennState}}, 
%Kazuhiro Shimasaku\altaffilmark{\ref{Tokyo}},
%Walter A. Siegmund\altaffilmark{\ref{UW}},
%Stephen Smee\altaffilmark{\ref{Maryland}},
%Yehuda Snir\altaffilmark{\ref{CarnegieMellon}},
%Chris Stoughton\altaffilmark{\ref{Fermilab}},
%Christopher Stubbs\altaffilmark{\ref{UW}},
%Alexander S.~Szalay\altaffilmark{\ref{JHU}},
%Gyula P.~Szokoly\altaffilmark{\ref{Potsdam}},
%Aniruddha R.~Thakar\altaffilmark{\ref{JHU}},
%Christy Tremonti\altaffilmark{\ref{JHU}},
%Douglas L. Tucker\altaffilmark{\ref{Fermilab}},
%Alan Uomoto\altaffilmark{\ref{JHU}},
%Dan vanden Berk\altaffilmark{\ref{Fermilab}},
%Michael S. Vogeley\altaffilmark{\ref{Drexel}},
%Patrick Waddell\altaffilmark{\ref{UW}},
%Brian Yanny\altaffilmark{\ref{Fermilab}},
%Naoki Yasuda\altaffilmark{\ref{NAOJ}},
%and Donald G.~York\altaffilmark{\ref{Chicago}}
}

\setcounter{address}{0}

@@

New York University, Center for Cosmology and Particle Physics, 4 Washington Place, New

@@

\setcounter{address}{1}
University of Toronto, Dept. of Computer Science, 6 King's College
Rd., Toronto, Ontario M5S 3G4, CANADA
\label{UToronto}}

@@

The method is designed to work well for a wide range of data sets.  It
uses photometry and spectroscopy of the Sloan Digital Sky Survey
(SDSS; \citealt{york00a}) the Galaxy Evolution Explorer (GALEX;
\citealt{martin05a}) in the ultraviolet, and the Two-Micron All Sky
Survey (2MASS; \citealt{skrutskie97a}) in the near infrared (NIR). In
addition, at higher redshifts, we use constraints from the Deep
Extragalactic Evolutionary Probe 2 (DEEP2; \citealt{davis03a,
faber03a}) and the Great Observatories Origins Deep Survey (GOODS;
\citealt{giavalisco04a}).  These and other data sets provide a huge
set of information about galaxy colors and spectra which we can use to
help understand their star-formation histories.

We note that some of the algorithmic techniques used here may have
applications in other areas of astrophysics. First, nonnegative matrix
factorization (NMF; \citealt{lee00a}), and the extensions to it we
describe here, is a particularly useful variant on principal component
analysis. Second, the nonnegative least squares algorithm of
\citet{sha02a} has the virtue of being extremely simple to implement.
These methods may find other applications in image analysis, spectroscopic
analysis, and model fitting in astrophysics.

We also release the templates in electronic form as well as an
implementation of the methods used to fit the templates to data. This
software {\tt kcorrect v4\_1} is distributed on the World Wide
Web{\footnote {\tt http://cosmo.nyu.edu/blanton/kcorrect/}}. It
consists of a core C library which performs most of the complex and
computationally intensive tasks, plus an IDL library that provides a
high-level interface.  This software is an update of two major earlier
releases ({\tt v1\_16}; \citealt{blanton03b}; {\tt v3\_2}). The
improvement over the previous version is twofold. First, we have
improved the templates such that they successfully fit galaxies in the
restframe UV, as observed by GALEX, DEEP2, and GOODS. Second, because
the templates are now completely model-based, the fits have a physical
interpretation in terms of a star-formation history. The IDL library
also has a number of new useful functions.

In Section \ref{algorithm}, we describe how to find suitable
templates, given a set of models and a set of data. We also describe
the data and models used here. In Section \ref{results}, we show the
results: the best-fit templates and how well they fit the data.  In
Section \ref{kcorrect}, we describe how to convert our results into an
estimate of the $K$-correction. In Section \ref{physical}, we describe
the physical interpretation of the templates and of fits to data.  In
Section \ref{sdss2bessell}, we present some simplified, linear
transformations between various bandpasses. In Section \ref{summary},
we summarize our results. Appendix \ref{nmf} describes the NMF
algorithm.  Appendix \ref{format} describes the electronic format of
our results.

Where necessary, we have assumed cosmological parameters $\Omega_0 =
0.3$, $\Omega_\Lambda = 0.7$, and $H_0 = 100$ $h$ km s$^{-1}$
Mpc$^{-1}$ (with $h=1$), unless otherwise noted. All magnitudes are
(unless otherwise noted) AB-relative.  Except where noted, I will
always be referring to the version of the {\tt kcorrect} software
labeled {\tt v4\_1\_4}.

@@

From the spectroscopic observations we know that galaxy spectra reside
in a low-dimensional subspace. Principal Component Analysis (PCA) of
galaxy spectra, introduced by \citet{connolly95b} and applied many
times since then, demonstrates that most of the variance in the
distribution of galaxies in spectral space can be explained using a
few templates. This means that, in the linear space of all possible
spectra, galaxies exist in only a small subspace. Therefore, even with
very heterogeneous data, we should be able to determine the properties
of this subspace.

Here we present an approach to combining heterogeneous data in order
to determine the properties of the subspace of galaxy spectra. Rather
than taking the model-free approach used by PCA, we here restrict the
space of possible spectra to those predicted from the high resolution
stellar population synthesis model of \citet{bruzual03a} and the
nebular emission line models of \citet{kewley01a}.  This approach both
constrains the problem appropriately and yields a natural theoretical
interpretation of the results in terms of star-formation histories.

In a nutshell, our algorithm does the following. Given the
observations (and uncertainties of those observations) available for
each galaxy, it finds the nonnegative linear combination of $N$
template star-formation histories which best predict those
observations in the $\chi^2$ sense. Given the entire set of galaxy
observations available, it also fits for the $N$ template
star-formation histories. The technical name for this algorithm is
Nonnegative Matrix Factorization (NMF; \citealt{lee99a, lee00a}), and
we describe it in detail in Appendix \ref{nmf}. Note that this problem
is not the same as the well-known nonnegative least squares problem,
which would fit for the best-fit nonnegative linear combination of
templates, but not for the form of the templates themselves.

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, the results yield, along with a
subspace in the space of all possible spectra, a natural
interpretation, which is the corresponding subspace in the space of
all possible star-formation histories. Second, it naturally handles
data uncertainties and missing data, which allows it to ignore that
variation which is due purely to statistical errors.  Third, it
handles the complications of observing galaxy spectra photometrically
using broad-band filters of galaxies at varying redshifts, whereas
standard PCA requires a constant grid of observations in rest-frame
wavelength space.

In the following subsections, we introduce the basis set of models we
use here (\S\ref{models}), the data we fit to (\S\ref{data}), and how
we combine the two to set up the NMF problem (\S\ref{setupnmf}).  In
Appendix \ref{nmf} we describe how to actually solve to NMF problem.

@@

Let us refer to these 485 basis templates as $M_{j}(\lambda)$,
expressed in units of ergs s$^{-1}$ \AA$^{-1}$.

We seek to reduce this full basis space to a subspace in which
galaxies actually exist. In particular, we seek five templates
$F_{i}(\lambda)$ built from nonnegative combinations of the
original basis set of $N$ templates:
\begin{equation}
F_{i}(\lambda) = \sum_j b_{ij} M_{j}(\lambda),
\end{equation}
in units of ergs s$^{-1}$ \AA$^{-1}$. In principle, we could seek any
number of templates; from our experiments, we have found that five
turns out to be a number large enough to explain the data we use
here. For each galaxy $k$ we want our model
${\hat{F}_\lambda}(\lambda)$ for their spectrum to be a nonegative sum
of these five templates:
\begin{equation}
{\hat{F}}_{k}(\lambda) = \sum_i a_{ki} F_{i}(\lambda).
\end{equation}

\subsection{The data}
\label{data}

@@

\subsection{Comparing data and models}
\label{setupnmf}

@@

For the spectra, we take the observed spectrum
$f_{k}(\lambda)$ (in ergs cm$^{-2}$ s$^{-1}$ \AA$^{-1}$) of
each galaxy $k$ at redshift $z$ (corresponding to a luminosity
distance $d_L(z)$ in cm; see \citealt{hogg99a}) and calculate the
restframe luminosity per unit wavelength:
\begin{equation}
{{F}}_{k}(\lambda) = 
{{f}}_{k}[\lambda (1+z)](1+z) (4 \pi d_L^2),
\end{equation}
That is, the spectrum is shifted due to the redshift, while the
integral of the numerator over wavelength (the total luminosity) is
constant with redshift, and the total flux is related to the total
luminosity by the inverse square law.  In addition, we smooth each
spectrum by such an amount that, given its estimated velocity
dispersion, its total velocity dispersion after smoothing is 300 km
s$^{-1}$ (there are a small number of galaxies with larger velocity
dispersions, which we leave unchanged).

@@

If one has a spectrum for a galaxy the
expression for the contribution to $\chi^2$ from each wavelength
$\lambda_l$ in the spectrum is then quite simple:
\begin{eqnarray}
\label{chi2spec}
\chi_{kl}^2 &=& \left[\frac{F_{k}(\lambda_l) -
{\hat{F}}_{k}(\lambda_l)} 
{\sigma^2_k(\lambda_l)}\right]^2 \cr
&=&
\left[\frac{F_{k}(\lambda_l) -
\sum_{ij} a_{ki} b_{ij} M_{jl}
}{\sigma^2_k(\lambda_l)}\right]^2. 
\end{eqnarray}

@@

Here we express the flux for each galaxy $k$ in units of AB maggies
$\mu_p$, which are defined as what one would measure in bandpass $p$
relative to the AB standard source. For example, if we transform our
spectral energy density basis function $M_{j}(\lambda)$ to a flux
density $m_{j}(\lambda)$, we can calculate the contribution of
that basis function to the predicted maggies as:
\begin{equation}
\label{maggies}
\mu_{jp} = 
\frac{\int_0^{\infty} d\lambda \lambda R_p(\lambda) {m}_{j}(\lambda)}
{\int_0^\infty d\lambda \lambda R_p(\lambda) f_{\mathrm{AB}}(\lambda) }
\end{equation}
Here, the response function $R_p(\lambda)$ is proportional to the
contribution to the detector signal of a photon with wavelength
$\lambda$ entering the Earth's atmosphere (or entering the telescope
for a space telescope). The AB standard source is $f_{
\mathrm{AB}}(\lambda) d\lambda = f_{
\mathrm{AB}} (\nu) d\nu$ and $f_{\mathrm{AB}}(\nu)=3631$ Jy $= 3.631 \times
10^{-20}$ erg s$^{-1}$ cm$^{-2}$ Hz$^{-1}$. Of course maggies $\mu$
are related to magnitudes $m$ as:
\begin{equation}
m = -2.5 \log_{10} \mu, 
\end{equation}
such that the AB standard source would (if it existed) have $\mu=1$
and $m=0$ for all bandpasses.

In this context it is worth noting that many authors
(e.g. \citealt{bessell90a}) tabulate the contribution to the detector
signal per {\it unit of energy} in photons of wavelength $\lambda$
instead of per {\it photon} with wavelength $\lambda$.  We will refer here
to the former quantity as $R_p'(\lambda)$, though in the literature it is
often referred to without a prime (and without any explicit
definition!). Clearly $R_p(\lambda) \propto R_p'(\lambda)/\lambda$,
since the higher the frequency of the photon, at a fixed response per
unit energy there is a higher response per photon. With this
substitution, one can reexpress Equation
\ref{maggies} appropriately in terms of $R_p'(\lambda)$.  
Generally, though not universally, authors tabulate $R_p'(\lambda)$
for bandpasses whose standards were originally calibrated using
energy-counting devices rather than photon-counting devices. However,
from the point of view of the analysis of the observations it is
irrelevant what the devices used for the standards and the
observations are, as long as one calculates $R_p(\lambda)$ and uses
Equation \ref{maggies}.

@@

The matrices $\eta_{jp}$ (related to the broad-band fluxes) and
$M_{jl}$ (related to the spectra) are totally fixed, and we combine
them into a single matrix $M_{jn}$ and the indices $p$ and $l$ into a
single index $n$, to handle both the broad-band fluxes and spectra
simultaneously. We can similarly combine our observations $\mu_{kp}$
and $F_{k}(\lambda_l)$ and their uncertainties $\sigma_{kp}$
and $\sigma_{k}(\lambda_l)$ into vectors $x_{kn}$ and $\sigma_{kn}$.
Then we can combine the Equations \ref{chi2spec} and
\ref{chi2photo} into a single equation:
\begin{equation}
\label{chi2}
\chi^2 = \sum_{kn} \left[ 
\frac{x_{kn} - \sum_{ij} a_{ki} b_{ij} M_{jn}}
{\sigma_{kn}} \right]^2
\end{equation}
There is a simple method that we describe in Appendix \ref{nmf} called
nonnegative matrix factorization (NMF) to iterate to the nonnegative
$a_{ki}$ and $b_{ij}$ which locally minimize Equation \ref{chi2}. The
basic method is implemented in a public piece of code named {\tt
NMF\_SPARSE} in the {\tt idlutils} distribution of IDL
utilities.\footnote{\tt http://skymaps.info}. As the name implies, our
implementation takes advantage of the fact that many of the matrix
operations are on very sparse matrices (for example, for each galaxy
with photometric data there are {\it no} spectroscopic data points and
{\it only} photometric data at a single redshift).

As we describe in Appendix \ref{nmf}, the NMF problem is not convex:
that is, there are multiple local minima. Our method finds one of the
local minima, but is not guaranteed to find the global minimum. Thus,
the reader who tries to reproduce our results using our methods or
others may (depending on their initial conditions) find different
local minima of $\chi^2$ than do we.
